---
title: "analysis_10Xdata"
author: "Agnes Paquet"
date: "21/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Note 1: Data files can be downloaded from Mendeley here:
https://data.mendeley.com/datasets/m289yp5skd/draft?a=65157631-161f-40a6-a718-23b0f9e6fa58
Data should remain private until this work is accepted for publication, please do not share without consent from the authors.

Note 2: this code was developped under R 3.6.1. It was not tested using more recent versions of R.

## Integration of CTRL, TRAIL_50mn and TRAIL 2H samples, and CTRl, TRAIL_50mn_dose1, TRAIL_50mn_dose2 

We are starting from the cell hashing data, after demultiplexing and cleaning. 
Data are available in a .rds file.

Preprocessing of all indiviudal samples:
- Extract the cell corresponding to each hashtag
- Set the stim parameter to the experimental condition
- Follow Seurat standard pipeline: 
+ Normalize to 10,000
+ Reduction of the number of features (vst, n=2,000 genes)

Final object is saved in a new .rds file.

Barcode correspondance 
A0251=CTRL
A0252=TRAIL 50mn, 25 ng
A0253=TRAIL 120mn, 10 ng
A0254=TRAIL 120mn, 25 ng
A0255=TRAIL 120mn, 40 ng

Please check paper for exact concentration values, as the data entered in the code might not be exact.

```{r}
library(Seurat)
library(pheatmap)
library(dplyr)
library(ggplot2)
library(cowplot)
 
library(ggrepel)
library(tidyverse)
library(fgsea)


```

```{r, eval=T}
## This part is now saved as a .rds file for speed


## Load object, export all samples
final <- readRDS("final_HTO_preprocess.rds")

ctrl <- subset(final, subset = HTO_classification=="A0251")
ctrl$stim <- "CTRL"
ctrl <- NormalizeData(ctrl, verbose = FALSE)
ctrl <- FindVariableFeatures(ctrl, selection.method = "vst", nfeatures = 2000)

t1 <- subset(final, subset = HTO_classification=="A0252")
t1$stim <- "TRAIL.50mn.d20"
t1 <- NormalizeData(t1, verbose = FALSE)
t1 <- FindVariableFeatures(t1, selection.method = "vst", nfeatures = 2000)

t2 <- subset(final, subset = HTO_classification=="A0253")
t2$stim <- "TRAIL.120mn.d10"
t2 <- NormalizeData(t2, verbose = FALSE)
t2 <- FindVariableFeatures(t2, selection.method = "vst", nfeatures = 2000)

t3 <- subset(final, subset = HTO_classification=="A0254")
t3$stim <- "TRAIL.120mn.d20"
t3 <- NormalizeData(t3, verbose = FALSE)
t3 <- FindVariableFeatures(t3, selection.method = "vst", nfeatures = 2000)

t4 <- subset(final, subset = HTO_classification=="A0255")
t4$stim <- "TRAIL.120mn.d40"
t4 <- NormalizeData(t4, verbose = FALSE)
t4 <- FindVariableFeatures(t4, selection.method = "vst", nfeatures = 2000)


## Integrate the data

#We use the Seurat integration function here. First thing we try is all 5 samples together.
#dims=20.
#It's quite long to run.
fib.anchors <- FindIntegrationAnchors(object.list = list(ctrl,t1,t2,t3,t4), dims = 1:20)
int <- IntegrateData(anchorset = fib.anchors, dims = 1:20)

## Run the analysis on the integrated dataset
# switch to integrated assay. The variable features of this assay are
# automatically set during IntegrateData
DefaultAssay(object = int) <- "integrated"

# Run the standard workflow for visualization and clustering
## First, we try without the cell cycle regression.
int <- ScaleData(object = int, verbose = FALSE, vars.to.regress = c("G2M.Score", "S.Score", "nCount_RNA"))

int <- RunPCA(object = int, npcs = 30, verbose = FALSE)
ElbowPlot(int) # we can pick 7 dims

int <- RunUMAP(object = int, reduction = "pca", dims = 1:7)
DimPlot(object = int, reduction = "umap", group.by = "stim")

int <- FindNeighbors(object = int, dims=1:7, do.plot=T)
int <- FindClusters(object = int, res=0.1)
DimPlot(object = int, reduction = "umap",label = T, label.size = 8)

tt =table(int@meta.data$integrated_snn_res.0.1, int$stim)
tt
round(tt*100/rowSums(tt),1)

VlnPlot(int,"nCount_RNA")
VlnPlot(int,"G2M.Score")
VlnPlot(int,"S.Score")

DimPlot(object = int, reduction = "umap",label = T, label.size = 8, split.by = 'stim')
int[["clusterV3"]] <- Idents(object = int)

saveRDS(int, file="integrated_cycle_corrected.rds")

```

## Reads processed data if needed
```{r}
int <- readRDS("integrated_cycle_corrected.rds")

```

```{r, fig.width=15,fig.height=9}

FeaturePlot(object = int, c("CDK1","UBE2C","TOP2A","CENPF", "HIST1H4C","MKI67"))

```

```{r}
FeaturePlot(object = int, c("LDHA","GAPDH"))

##Idents(object = int) <- int$clusterV3

VlnPlot(object = int, c("CDK1","UBE2C","TOP2A","CENPF", "HIST1H4C"))

  
```



# GSEA on the markers of cluster 2 and 3
Note: we find mostly cell cyle genesm mitotic_spindle genes, and metabolism/glycolysis genes.
Note: not run 
```{r, eval=F}
markers = FindAllMarkers(int)

library(fgsea)
library(tidyverse)

## Note the Hallmark Pathway file and KEGG pathways must be downloaded elsewhere
pathways.hallmark <- gmtPathways("h.all.v6.2.symbols.gmt.txt")
pathways.kegg <- gmtPathways("c2.cp.kegg.v6.2.symbols.gmt.txt")

## MA-plot from average matrix
c2.mark = markers[markers$cluster==2,]
c2.mark

ranki <- deframe(c2.mark[,c("gene","avg_logFC")])
print(head(ranki))
fres <- fgsea(pathways=pathways.hallmark, stats=ranki, nperm=1000)
fres <- fres %>%
  as_tibble() %>%
  arrange(desc(NES))
print(fres)

ggplot(fres, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA, cluster 2") + 
  theme_minimal()

c3.mark = markers[markers$cluster==3,]
c3.mark

ranki <- deframe(c3.mark[,c("gene","avg_logFC")])
print(head(ranki))
fres <- fgsea(pathways=pathways.hallmark, stats=ranki, nperm=1000)
fres <- fres %>%
  as_tibble() %>%
  arrange(desc(NES))
print(fres)

ggplot(fres, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA, cluster 3") + 
  theme_minimal()


ranki <- deframe(c3.mark[,c("gene","avg_logFC")])
print(head(ranki))
fres <- fgsea(pathways=pathways.kegg, stats=ranki, nperm=1000)
fres <- fres %>%
  as_tibble() %>%
  arrange(desc(NES))
print(fres)

fres = fres[fres$pval<0.1,]
ggplot(fres, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.05)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Top KEGG pathways NES from GSEA, cluster 3") + 
  theme_minimal()



```


## Set the groups to the experimental condition
# Then run the differential analysis for the comp of interest

```{r}
Idents(object = int) <- int$stim
levels(int)

deg1 = FindMarkers(int, ident.1 = "TRAIL.50mn.d20", ident.2 = "CTRL", min.pct = 0.5, logfc.threshold = 0.1)
deg1

deg2 = FindMarkers(int, ident.1 = "TRAIL.120mn.d20", ident.2 = "CTRL", min.pct = 0.5, logfc.threshold = 0.1)
deg2

deg3 = FindMarkers(int, ident.1 = "TRAIL.120mn.d40", ident.2 = "CTRL", min.pct = 0.5, logfc.threshold = 0.1)
deg3

```



##  UMAP CTRL + 50mnD20 + 120mnD20
```{r}
Idents(object = int) <- int$stim
levels(int)
sub1 = subset(int, idents = c("CTRL","TRAIL.50mn.d20","TRAIL.120mn.d20"))
sub1$stim <- factor(sub1$stim, levels=c("CTRL","TRAIL.50mn.d20","TRAIL.120mn.d20"))
sub1[["clusterV3"]] <- Idents(object = sub1)

sub = subset(int, idents = c("CTRL","TRAIL.50mn.d20"))

DimPlot(sub1,pt.size=2)
DimPlot(sub1,pt.size=2, split.by = "stim")

## PDF UMAPs by exp condition
pdf("UMAP_CTRL_TR50D20_TR120D20_byStim.pdf", paper="a4")
DimPlot(sub1,pt.size=2)
DimPlot(sub1,pt.size=2, split.by = "stim")
dev.off()

DimPlot(sub,pt.size=2)
DimPlot(sub,pt.size=2, split.by = "stim")

pdf("UMAP_CTRL_TR50D20_byStim.pdf", paper="a4")
DimPlot(sub,pt.size=2)
DimPlot(sub,pt.size=2, split.by = "stim")
dev.off()

## PDF UMAPs by cluster

Idents(object = sub) <- int$clusterV3
Idents(object = sub1) <- int$clusterV3

sub1$stim <- factor(sub1$stim, levels=c("CTRL","TRAIL.50mn.d20","TRAIL.120mn.d20"))

DimPlot(sub1,pt.size=2)
DimPlot(sub1,pt.size=2, split.by = "stim")

pdf("UMAP_CTRL_TR50D20_TR120D20_byStim_withClusters.pdf", paper="a4")
DimPlot(sub1,pt.size=2)
DimPlot(sub1,pt.size=2, split.by = "stim")
dev.off()

DimPlot(sub,pt.size=2)
DimPlot(sub,pt.size=2, split.by = "stim")

pdf("UMAP_CTRL_TR50D20_byStim_withClusters.pdf", paper="a4")
DimPlot(sub,pt.size=2)
DimPlot(sub,pt.size=2, split.by = "stim")
dev.off()


```

##Markers and UMAP for each sample
```{r}
s1 =  subset(int, subset = stim=="CTRL") 
Idents(s1) = "clusterV3"
m1=FindAllMarkers(s1)
top5 <- m1 %>% group_by(cluster) %>% top_n(5, avg_logFC)
top10 <- m1 %>% group_by(cluster) %>% top_n(10, avg_logFC)
top20 <- m1 %>% group_by(cluster) %>% top_n(20, avg_logFC)

table(m1$cluster)

library(gplots)

DoHeatmap(object = s1, features = top10$gene)

pdf("Heatmap_CTRL_allcells.pdf", width=10,height=8, useDingbats = F)
DoHeatmap(object = s1, features = top10$gene, raster=F)
dev.off()

DoHeatmap(object = s1, features = top10$gene) + scale_fill_gradientn(colors = c("blue", "white", "red"))

pdf("Heatmap_CTRL_allcells_redblue.pdf", width=10,height=8, useDingbats = F)
DoHeatmap(object = s1, features = top10$gene, raster=F) + scale_fill_gradientn(colors = c("blue", "white", "red"))
dev.off()



```



```{r}
s2 =  subset(int, subset = stim=="TRAIL.50mn.d20") 
Idents(s2) = "clusterV3"
m2=FindAllMarkers(s2)
top10 <- m2 %>% group_by(cluster) %>% top_n(10, avg_logFC)

table(m2$cluster)

DoHeatmap(object = s2, features = top10$gene)

pdf("Heatmap_TR50mnD20_allcells.pdf", width=10,height=8, useDingbats = F)
DoHeatmap(object = s2, features = top10$gene, raster=F)
dev.off()

DimPlot(s2, pt.size=2)

DoHeatmap(object = s2, features = top10$gene) + scale_fill_gradientn(colors = c("blue", "white", "red"))

pdf("Heatmap_TR50mnD20_allcells_redblue.pdf", width=10,height=8, useDingbats = F)
DoHeatmap(object = s2, features = top10$gene, raster=F) + scale_fill_gradientn(colors = c("blue", "white", "red"))
dev.off()

write.table(m2, file="ClusterMarkers-TR50mnD20_V3.txt", sep="\t",quote=F,row.names=T)
```


```{r}
s3 =  subset(int, subset = stim=="TRAIL.120mn.d20") 
Idents(s3) = "clusterV3"
m3=FindAllMarkers(s3)
top10 <- m3 %>% group_by(cluster) %>% top_n(10, avg_logFC)

table(m3$cluster)

DoHeatmap(object = s3, features = top10$gene)

pdf("Heatmap_TR120mnD20_allcells.pdf", width=12,height=10)
DoHeatmap(object = s3, features = top10$gene)
dev.off()

DoHeatmap(object = s3, features = top10$gene) + scale_fill_gradientn(colors = c("blue", "white", "red"))

pdf("Heatmap_TR120mnD20_allcells_redblue.pdf", width=12,height=10)
DoHeatmap(object = s3, features = top10$gene) + scale_fill_gradientn(colors = c("blue", "white", "red"))
dev.off()

write.table(m3, file="ClusterMarkers-TR50mnD20_V3.txt", sep="\t",quote=F,row.names=T)
```

## markers, CTRL + 50mn together
```{r}
sub = subset(int, subset=stim %in% c("CTRL","TRAIL.50mn.d20"))
Idents(sub) = "clusterV3"

mm = FindAllMarkers(sub)
top10 <- mm %>% group_by(cluster) %>% top_n(10, avg_logFC)
top20 <- mm %>% group_by(cluster) %>% top_n(20, avg_logFC)

table(mm$cluster)

cluster.averages <- AverageExpression(object = sub, return.seurat = TRUE, add.ident = "stim",assay="RNA" )

DoHeatmap(object = cluster.averages, features = top20$gene)

pdf("Heatmap_CTRL_TR50mnD20_averages.pdf", width=8,height=8, useDingbats = F)
DoHeatmap(object = cluster.averages, features = top20$gene, raster=F)
dev.off()


pdf("Heatmap_CTRL_TR50mnD20_average_redblue.pdf", width=8,height=12)
DoHeatmap(object = sub, features = top20$gene) #+ scale_fill_gradientn(colors = c("blue", "white", "red"))
dev.off()


write.table(mm, file="ClusterMarkers-CTRL_TR50mnD20_V3.txt", sep="\t",quote=F,row.names=T)

```

## GSEA analysis for the cluster markers
```{r, fig.height=10}
library(fgsea)
library(tidyverse)
pathways.hallmark <- gmtPathways("h.all.v6.2.symbols.gmt.txt")
pathways.kegg <- gmtPathways("c2.cp.kegg.v6.2.symbols.gmt.txt")

rankc0 <- deframe(mm[mm$cluster==0,c("gene","avg_logFC")])
print(head(rankc0))
fres <- fgsea(pathways=pathways.hallmark, stats=rankc0, nperm=1000)
fres <- fres %>%
  as_tibble() %>%
  arrange(desc(NES))
print(fres)

ggplot(fres, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA, cluster 0 vs other") + 
  theme_minimal()


rankc1 <- deframe(mm[mm$cluster==1,c("gene","avg_logFC")])
print(head(rankc1))
fres <- fgsea(pathways=pathways.hallmark, stats=rankc1, nperm=1000)
fres <- fres %>%
  as_tibble() %>%
  arrange(desc(NES))
print(fres)

ggplot(fres, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA, cluster 1 vs other") + 
  theme_minimal()


fres <- fgsea(pathways=pathways.kegg, stats=rankc1, nperm=1000)
fres <- fres %>%
  as_tibble() %>%
  arrange(desc(NES))
print(fres)

ggplot(fres[fres$pval<0.1,], aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA, cluster 1 vs other") + 
  theme_minimal()


rankc2 <- deframe(mm[mm$cluster==2,c("gene","avg_logFC")])
print(head(rankc2))
fres <- fgsea(pathways=pathways.hallmark, stats=rankc2, nperm=1000)
fres <- fres %>%
  as_tibble() %>%
  arrange(desc(NES))
print(fres)

ggplot(fres, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA, cluster 2 vs other") + 
  theme_minimal()


fres <- fgsea(pathways=pathways.kegg, stats=rankc2, nperm=1000)
fres <- fres %>%
  as_tibble() %>%
  arrange(desc(NES))
print(fres)

ggplot(fres[fres$pval<0.1,], aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA, cluster 1 vs other") + 
  theme_minimal()

```



## GSEA analysis for the cluster markers. CTRL sample alone
```{r, fig.height=10}
library(fgsea)
library(tidyverse)
pathways.hallmark <- gmtPathways("h.all.v6.2.symbols.gmt.txt")
pathways.kegg <- gmtPathways("c2.cp.kegg.v6.2.symbols.gmt.txt")

rankc0 <- deframe(m1[m1$cluster==0,c("gene","avg_logFC")])
print(head(rankc0))
fres <- fgsea(pathways=pathways.hallmark, stats=rankc0, nperm=1000)
fres <- fres %>%
  as_tibble() %>%
  arrange(desc(NES))
print(fres)

ggplot(fres, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA, cluster 0 vs other") + 
  theme_minimal()


rankc1 <- deframe(m1[m1$cluster==1,c("gene","avg_logFC")])
print(head(rankc1))
fres <- fgsea(pathways=pathways.hallmark, stats=rankc1, nperm=1000)
fres <- fres %>%
  as_tibble() %>%
  arrange(desc(NES))
print(fres)

ggplot(fres, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA, cluster 1 vs other") + 
  theme_minimal()


rankc2 <- deframe(m1[m1$cluster==2,c("gene","avg_logFC")])
print(head(rankc2))
fres <- fgsea(pathways=pathways.hallmark, stats=rankc2, nperm=1000)
fres <- fres %>%
  as_tibble() %>%
  arrange(desc(NES))
print(fres)

ggplot(fres, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA, cluster 2 vs other") + 
  theme_minimal()



```






## GSEA analysis for the cluster markers. TRAIL T50 D20 sample alone
```{r, fig.height=10}

rankc0 <- deframe(m2[m2$cluster==0,c("gene","avg_logFC")])
print(head(rankc0))
fres <- fgsea(pathways=pathways.hallmark, stats=rankc0, nperm=1000)
fres <- fres %>%
  as_tibble() %>%
  arrange(desc(NES))
print(fres)

ggplot(fres, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA, cluster 0 vs other") + 
  theme_minimal()


rankc1 <- deframe(m2[m2$cluster==1,c("gene","avg_logFC")])
print(head(rankc1))
fres <- fgsea(pathways=pathways.hallmark, stats=rankc1, nperm=1000)
fres <- fres %>%
  as_tibble() %>%
  arrange(desc(NES))
print(fres)

ggplot(fres, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA, cluster 1 vs other") + 
  theme_minimal()


rankc2 <- deframe(m2[m2$cluster==2,c("gene","avg_logFC")])
print(head(rankc2))
fres <- fgsea(pathways=pathways.hallmark, stats=rankc2, nperm=1000)
fres <- fres %>%
  as_tibble() %>%
  arrange(desc(NES))
print(fres)

ggplot(fres, aes(reorder(pathway, NES), NES)) +
  geom_col(aes(fill=padj<0.1)) +
  coord_flip() +
  labs(x="Pathway", y="Normalized Enrichment Score",
       title="Hallmark pathways NES from GSEA, cluster 2 vs other") + 
  theme_minimal()



```

