---
title: "Figures_Roux_ms_fate-seq_file_1"
author: "Agnes Paquet"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width=7,fig.height = 5)
```

```{r}
# load required package
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(gplots))
suppressPackageStartupMessages(library(pheatmap))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggrepel))
suppressPackageStartupMessages(library(plyr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(SingleCellExperiment))
suppressPackageStartupMessages(library(scater))
suppressPackageStartupMessages(library(scran))
suppressPackageStartupMessages(library(factoextra))
suppressPackageStartupMessages(library(FactoMineR))


```

## Heatmap based on HVG

Preprocessing using scran. HVG selected using Lun method, regardless of cell labels 

Based on: https://bioconductor.org/packages/release/workflows/vignettes/simpleSingleCell/inst/doc/xtra-3-var.html


```{r}

setwd("C:/Users/Kevin/Projets/Collaborations/Roux/Figure_032019/Figures_01292020/")
sc <- read.delim("GEO_RawCounts_LCMcells.txt", stringsAsFactors = F, row.names=1)
serie <- read.delim(file = "serie.FinalN19_GEO_extended.txt", stringsAsFactors = F)

## check that serie and count data are ordered the same
ix <- match(serie$final.names,colnames(sc))
## check cbind(serie$final.names,colnames(sc)[ix])

sc <- sc[,ix] ## reorder the columns
rm(ix);gc(T)

pheno <- factor(serie$pheno, levels = c("Low","High"))
names(pheno) = colnames(sc)
ercc = grep("ERCC-", rownames(sc))

mycols = brewer.pal(8,"Set1")

annotation_col = data.frame(Pheno=serie$pheno,
                            Batch = as.numeric(as.factor(serie$Sample.Date)))
rownames(annotation_col) = serie$final.names

ann_colors = list(Pheno = c(Low=mycols[2], High=mycols[1]),
                  Batch=rainbow(8))

```

## Data normalization using scnorm/scran

```{r}

sce.hsc <- SingleCellExperiment(list(counts=as.matrix(sc)))
dim(sce.hsc)

is.spike <- grepl("^ERCC-", rownames(sce.hsc))
isSpike(sce.hsc, "ERCC") <- is.spike
summary(is.spike)

to.keep <- nexprs(sce.hsc, byrow=TRUE) > 0
sce.hsc <- sce.hsc[to.keep,]
summary(to.keep)

sce.hsc <- computeSumFactors(sce.hsc,min.mean=0.1, sizes=seq(3,5))
summary(sizeFactors(sce.hsc))

sce.hsc <- computeSpikeFactors(sce.hsc, type="ERCC", general.use=FALSE)
summary(sizeFactors(sce.hsc, "ERCC"))

sce.hsc <- normalize(sce.hsc)

var.fit <- trendVar(sce.hsc, parametric=TRUE, loess.args=list(span=0.4))
#var.out <- decomposeVar(sce.hsc, var.fit)
var.out <- decomposeVar(sce.hsc, var.fit)

## Plot the HVG curve
plot(var.out$mean, var.out$total, pch=16, cex=0.6, xlab="Mean log-expression", 
    ylab="Variance of log-expression")
curve(var.fit$trend(x), col="dodgerblue", lwd=2, add=TRUE)
cur.spike <- isSpike(sce.hsc)
points(var.out$mean[cur.spike], var.out$total[cur.spike], col="red", pch=16)

hvg.out <- var.out[which(var.out$p.value <= 0.05),]

```

## Normalized data for exploratory analysis

Based on scran normalization, no ERCC

```{r,fig.width=8,fig.height=10}
norm <- logcounts(sce.hsc)
ercc <- grep("ERCC-",rownames(norm))
## Heatmap of normalized counts for HVG

annotation_col = data.frame(Pheno=serie$pheno)
rownames(annotation_col) = serie$final.names

ann_colors = list(Pheno = c(Low=mycols[2], High=mycols[1]))

hvg <- rownames(var.out[which(var.out$FDR <= 0.05),])

pheatmap(norm[hvg,],scale="row",
          annotation_col=annotation_col,
          annotation_colors = ann_colors,
          cluster_distance_cols = "correlation",
          clustering_method = "ward.D",
          fontsize_row=10,fontsize_col = 18)

```


## Cell cycle analysis

Note: R code based on a blog post. Original code is available here:
https://jdblischak.github.io/singleCellSeq/analysis/cell-cycle.html

We used our own set of genes for cell cycle assessment.

## Cell cycle assignment
```{r, eval=T}
cell_cycle_genes <- read.delim("Barbry-CellCycle_GeneSets.txt", stringsAsFactors = F)

## create 5 lists of 5 phases (de-level and then remove "")
cell_cycle_genes_list <- apply(cell_cycle_genes,2, 
                               function(x){ temp <- gsub(" ","",x);return(temp[temp!=""])})

## create phase score for each cell
phase.score = c()
for(i in 1:5)
{
  g1.s = cell_cycle_genes_list[[i]]
  tmp.g1s = norm[rownames(norm) %in% g1.s,]
  tmp.g1s = rbind(tmp.g1s, mean=colMeans(tmp.g1s))

  ### use transpose to compute cor matrix
  cor_matrix <- cor(t(tmp.g1s))

  #### take the numbers for correlation of each genes with mean values for the set
  cor_vector <- cor_matrix[,dim(cor_matrix)[1]]
  #### restrict to correlation >= 0.3
  tmp.g1s.filter <- norm[rownames(norm) %in% names(cor_vector[cor_vector >= 0.3]),]
  g1s.score = colMeans(tmp.g1s.filter)
  phase.score = cbind(phase.score, g1s.score)
  rm(g1s.score, g1.s, tmp.g1s, tmp.g1s.filter)
}
colnames(phase.score) = names(cell_cycle_genes_list)

## Normalization of score for each phase

flexible_normalization <- function(data_in,by_row=TRUE){
  if(by_row){
    row_mean <- apply(data_in,1,mean)
    row_sd   <- apply(data_in,1,sd)
    output <- data_in
    for(i in 1:dim(data_in)[1]){
      output[i,] <- (data_in[i,] - row_mean[i])/row_sd[i]
    }
  }
  #### if by column
  if(!by_row){
    col_mean <- apply(data_in,2,mean)
    col_sd   <- apply(data_in,2,sd)
    output <- data_in
    for(i in 1:dim(data_in)[2]){
      output[,i] <- (data_in[,i] - col_mean[i])/col_sd[i]
    }
  }
  output
}

#### apply the normalization function
## first normalized for each phase
phase.score.norm1 <- flexible_normalization(phase.score,by_row=FALSE)
## then normalized of each cell
phase.score.norm2 <- flexible_normalization(phase.score.norm1,by_row=TRUE)

#####################################################
## Assign phase to each cell
tmp = data.frame(Name= rownames(phase.score.norm2), phase.score.norm2)
cell_phase <- apply(phase.score.norm2,1,function(x) {colnames(phase.score.norm2)[which.max(x)]})
tmp$cell_phase = cell_phase

ord.mat=rbind(dplyr::arrange(as.data.frame(tmp[cell_phase=="G1.S",]),desc(G1.S)),
              dplyr::arrange(as.data.frame(tmp[cell_phase=="S",]),desc(S)),
              dplyr::arrange(as.data.frame(tmp[cell_phase=="G2.M",]),desc(G2.M)),
              dplyr::arrange(as.data.frame(tmp[cell_phase=="M",]),desc(M)),
              dplyr::arrange(as.data.frame(tmp[cell_phase=="M.G1",]),desc(M.G1)))

ord.mat$cell_phase = factor(ord.mat$cell_phase, levels=c("G1.S", "S", "G2.M","M", "M.G1"))

## Heatmap of cell cycle scores with batch and phenotype info
row.names(ord.mat) <- ord.mat$Name

annotation_col = data.frame(Pheno=serie$pheno,
                            Batch = as.factor(as.numeric(as.factor(serie$Sample.Date))))
rownames(annotation_col) = serie$final.names

ann_colors = list(Pheno = c(Low=mycols[2], High=mycols[1]))

br = bluered(11)

pheatmap(ord.mat[,2:6], annotation_row = annotation_col,
         col=br, annotation_colors = ann_colors,
         cluster_rows = F, cluster_cols = F, border_color = "black")


```


## PCA and correlation analysis

```{r}

norm <- logcounts(sce.hsc)
ercc <- grep("ERCC-",rownames(norm))

res.pca <- PCA(t(norm[-ercc,]),scale.unit=T, graph = F)

tmp = data.frame(id = rownames(res.pca$ind$coord),
                 res.pca$ind$coord,
                 pheno=pheno)

gg <- ggplot(tmp, aes(x=Dim.1,y=Dim.2,color=pheno)) + 
         geom_point(size=3) +
         stat_ellipse(type="norm") + 
         theme_classic() + 
         geom_text_repel(aes(label=id), size=3, 
                        show.legend=FALSE, force=2) + 
      theme(legend.position="none") + scale_color_manual(values = mycols[c(2,1)])
gg + xlab("Dim 1") + ylab("Dim 2")

```



## Predicted trail-response signature
 
aka differential expression, high vs low
 
Differential analysis is performed using edgeR

```{r}
## prepare design matrix
ercc = grep("ERCC-",rownames(sc))
dge <- DGEList(counts=sc[-ercc,], group=serie$pheno)
keep <- rowSums(sc[-ercc,]>5) > 5
dge <- dge[keep,]
#dim(dge)# 3095

dge <- calcNormFactors(dge)
serie$pheno = factor(serie$pheno, levels=c("Low","High"))
design = model.matrix(~serie$pheno)
#design
colnames(design) = c("Int", "HighvsLow")

## edgeR robust
dge <- estimateGLMRobustDisp(dge, design = design)
fit <- glmFit(dge, design = design) 
lrt <- glmLRT(fit) 
tt <- topTags(lrt, n = Inf) 

#write.table(tt$table, file="Results_edgeR_Robust_GLM_FINAL_2020.xls", sep="\t",quote=F,col.names=NA) 

res = tt$table
```
 
```{r}
res = tt$table
deg <- which(res$FDR<0.05)

res$Geneid = rownames(res)
mygenes <- sort(c( "PRSS56", "SIVA1", "SLC25A1", "DNM1L","C11orf83", "UBE2D4","BLOC1S1"))

## set-up various filter for plotting labels
res$toplot <- ifelse(res$FDR<0.05 ,1,0)
res$toplot[res$Geneid %in% mygenes] <- 0

res$target <- rep(0,nrow(res))
res[mygenes,"target"] <- 1

res$toplot2 <- ifelse(res$FDR<0.1 | res$Geneid %in% mygenes,1,0)

gg <- ggplot(data=res, aes(x=logCPM, y=logFC)) +
      geom_point(colour="grey") +
      geom_point(data=subset(res,target==1),colour = "red", size=2) +
      geom_text_repel(data=subset(res,target==1), aes(label=Geneid), size=3, 
                      show.legend=FALSE, force=2) + 
      theme_classic() +
      theme(legend.position="none")
gg + ggtitle("MA-plot: High vs Low")

gg <- ggplot(data=res, aes(x=logFC, y=-log10(FDR))) +  
      geom_point(color="grey") +
      geom_point(data=subset(res,target==1),colour = "red", size=2) +
      geom_text_repel(data=subset(res,target==1), aes(label=Geneid), size=3, 
                      show.legend=FALSE, force=2) + 
      theme_classic() +
      theme(legend.position="none")
gg + ggtitle("Volcano-plot: High vs Low")

```


+ Heatmap of Differentially Expressed Genes


```{r, fig.width=10, fig.height=12}
res$toplot2 <- ifelse(res$FDR<0.1,1,0)

annotation_col = data.frame(Pheno=serie$pheno)
rownames(annotation_col) = serie$final.names

ann_colors = list(Pheno = c(Low=mycols[2], High=mycols[1]))

color.palette = colorRampPalette(c("midnightblue","dodgerblue3","white","goldenrod1","darkorange2"), space="Lab")(15)

pheatmap(norm[res$Geneid[res$toplot2==1],],
         scale="row",border_color = NA,
         clustering_method = "ward.D",
         cutree_cols = 2, cutree_rows = 3,
         col=color.palette,
         annotation_col = annotation_col, annotation_colors = ann_colors)

```

## sessionInfo
```{r}
sessionInfo()
```

